Seasonality in governing upper-ocean dynamics at submesoscales

### 2016-06-04 writeup/rocha_etal.tex
Fixes kenged
Update subsection heads

Using the output of a $1/24^\circ$ global simulation, we show that the near-surface
expression of inertia-gravity waves undergoes a strong seasonal cycle that is out-of-phase
with the seasonal cycle of the balanced variability. This seasonal cycle is associated
with changes in the upper-ocean
stratification. Consistent with previous studies,
deep late Winter mixed-layers are prone to
shallow baroclinic instabilities that are roughly in geostrophic balance
and flux energy upscale \citep{sasaki_etal2014,callies_etal2016},
driving a mild seasonal modulation of the mesoscales \citep{sasaki_etal2014,qiu_etal2014}.
The horizontally divergent part of the flow, however, peaks in late Summer/early Fall,
when the upper-ocean is strongly stratified. Hence these results suggest a
strong seasonal modulation of the dominant upper-ocean submesoscale dynamics:
quasi-balanced turbulence in late Winter/early Spring and
quasi-linear inertia-gravity waves in late Summer/early Fall.

\section{The LLC2160 model output}
We use the output of a latitude-longitude polar cap (LLC)
MITgcm  numerical simulation  \citep{forget_etal2015}. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and 6-hourly surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years from March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution  of the
LLC2160 simulation is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.
 (see figure \ref{fig3}).

@@

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows are supra-inertial and project
on scales smaller than 100 km: daily-averaging or smoothing the velocity fields
 (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Interestingly, the seasonal cycle in horizontal divergence is out-of-phase with the
seasonal cycle of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\hspace{-1.25cm}\includegraphics[width=.7\textwidth]{figs/fig1_1.pdf}\\
\vspace{-.125cm}
\includegraphics[width=.9\textwidth]{figs/fig1_2.pdf}
 \caption{(a) The study region with the subregion where the LLC outputs are
          analyzed. Snapshots of surface vorticity (b and c) and a transect
          of potential density at 165$^\circ$E (d and e). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.9\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square of vorticity (a),
  horizontal divergence (b), and stratification in the upper 50 m (c).
  \textcolor{red}{May add a time series of baroclinic production $\overline{w'b'}$ in
  the mixed layer. }.}
  \label{fig2}
  \end{center}
\end{figure*}

>>>>

Using the output of two new primitive equation $1/24^\circ$ and $1/24^\circ$ global
 simulations with embedded tides, we show that the near-surface
expression of inertia-gravity waves undergoes a strong seasonal cycle that is out-of-phase
with the seasonal cycle of the balanced variability. This seasonal cycle is associated
with changes in the upper-ocean
stratification. Consistent with previous studies,
deep late Winter mixed-layers are prone to
shallow baroclinic instabilities that are roughly in geostrophic balance
and flux energy upscale \citep{sasaki_etal2014,callies_etal2016},
driving a mild seasonal modulation of the mesoscales \citep{sasaki_etal2014,qiu_etal2014}.
The horizontally divergent part of the flow, however, peaks in late Summer/early Fall,
when the upper-ocean is strongly stratified. Hence these results suggest a
strong seasonal modulation of the dominant upper-ocean submesoscale dynamics:
quasi-balanced turbulence in late Winter/early Spring and
quasi-linear inertia-gravity waves in late Summer/early Fall.

\section{The LLC model outputs}
We use the output of a latitude-longitude polar cap (LLC)
MITgcm  numerical simulation  \citep{forget_etal2015}. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and 6-hourly surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years from March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution  of the
LLC2160 simulation is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.
 (see figure \ref{fig3}).

@@

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows are supra-inertial and project
on scales smaller than 100 km: daily-averaging or smoothing the velocity fields
 (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Interestingly, the seasonal cycle in horizontal divergence is out-of-phase with the
seasonal cycle of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\hspace{-1.25cm}\includegraphics[width=.7\textwidth]{figs/fig1_1.pdf}\\
\vspace{-.125cm}
\includegraphics[width=.9\textwidth]{figs/fig1_2.pdf}
 \caption{(a) The study region with the subregion where the LLC outputs are
          analyzed. LLC 4320 (1/48$^\circ$) snapshots of surface vorticity (b and c) and a transect
          of potential density at 165$^\circ$E (d and e). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.9\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square (RMS) of vorticity (a) and
  horizontal divergence (b) in the LLC outputs and gridded AVISO data.}
  \label{fig2}
  \end{center}
\end{figure*}
